---
title: "Tufte Visualization Principles: Practice Exercises"
author: "R Data Visualization Practice"
format:
  html:
    toc: true
    code-fold: false
    theme: cosmo
editor: visual
---

## Introduction

This document provides practice exercises in creating data visualizations that adhere to Edward Tufte's principles of effective graphical design. Tufte's principles emphasize:

1.  **Data-ink ratio**: Maximize the proportion of ink devoted to displaying actual data
2.  **Minimize chartjunk**: Remove unnecessary decorative elements
3.  **Show data variation, not design variation**: Let the data speak for itself
4.  **Use small multiples**: Facilitate comparison through repetition
5.  **Integrate text and graphics**: Labels should be close to the data they describe
6.  **Avoid misleading representations**: Use appropriate scales and proportions

## Dataset Description

We'll be working with a sales dataset (`sample_data.csv`) containing:

- **year**: Year of observation (2020-2023)
- **region**: Geographic region (North, South, East, West)
- **product**: Product category (Product_A, Product_B, Product_C)
- **revenue**: Sales revenue in dollars
- **units_sold**: Number of units sold
- **customer_satisfaction**: Customer satisfaction score (1-5 scale)
- **market_share**: Market share percentage

## Setup

```{r setup, message=FALSE, warning=FALSE}
# Load required libraries
library(tidyverse)
library(ggplot2)
library(ggthemes)

# Read the data
data <- read_csv("sample_data.csv")

# Preview the data structure
glimpse(data)
```

------------------------------------------------------------------------

## Exercise 1: High Data-Ink Ratio Line Chart

**Task**: Create a line chart showing revenue trends over time for each product, applying Tufte's principle of maximizing the data-ink ratio.

**Requirements**: - Remove unnecessary grid lines - Minimize or remove axis lines where appropriate - Use direct labels instead of a legend - Remove background shading - Keep only essential elements

```{r exercise-1}
# YOUR CODE HERE
data |>
  mutate(region = factor(region, levels = c("North", "East", "South", "West"))) |>
  ggplot(aes(x = year, y = revenue, color = product)) +
  geom_line() +
  facet_wrap(~ region) +
labs(x = "Year", y="Revenue (USD)") +
theme(
  panel.grid        = element_blank(),
  panel.border      = element_blank(),
  axis.line         = element_line(colour = "black", linewidth = 0.3),
  panel.background  = element_blank()
) +
  geom_text(
    data = data |> filter(year == max(year)),  # last point of each line
    aes(label = product),
    hjust = -0.1,    # nudge right of the line end
    size = 3.5
  ) +
  theme(legend.position = "none") +
  coord_cartesian(clip = "off") +  # allow labels to render outside plot area
  theme(plot.margin = margin(r = 60))  # make room on the right for labels

```

**Expected Output**: A clean line chart with minimal non-data ink, where product lines are clearly labeled and trends are easy to identify.

------------------------------------------------------------------------

## Exercise 2: Minimal Bar Chart

**Task**: Create a bar chart comparing total revenue by region for 2023, following minimalist design principles.

**Requirements**: - Remove grid lines - Remove chart borders - Use subtle colors or grayscale - Label bars directly (no legend needed) - Remove unnecessary axis elements

```{r exercise-2}
# YOUR CODE HERE
data <- data |> mutate(product = gsub("_", " ", product))
data |> 
  filter(year == "2023") |> 
  ggplot(aes(x = product, y = revenue)) + 

  geom_col(fill = "lightgrey") + 
  geom_hline(yintercept = seq(0, max(data$revenue), by = 50000),
             colour = "white", linewidth = 0.1) +
  facet_wrap(~ region, labeller = labeller(region = toupper)) +
  theme(
    panel.grid       = element_blank(),
    panel.border     = element_blank(),
    axis.line        = element_line(colour = "black", linewidth = 0.3),
    panel.background = element_blank(),
    axis.line.x      = element_blank(),
    axis.ticks.x     = element_blank(),
    axis.line.y      = element_blank(),
    axis.ticks.y = element_blank(),
    axis.text.y  = element_blank(),
  axis.title.y = element_blank(),
    strip.background = element_blank(),
    legend.position  = "none"
  ) +
theme(strip.text = element_text(size = 11)) +
scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
geom_text(aes(label = scales::comma(revenue)), vjust = -0.5, size = 3) +
  labs(x = "", y = "Revenue (USD)", title = "Product C generates most revenue in all regions in 2023")
```

**Expected Output**: A clean, uncluttered bar chart that focuses attention on the data values.

------------------------------------------------------------------------

## Exercise 3: Small Multiples for Comparison

**Task**: Create small multiples (faceted plots) to compare revenue trends across all four regions over time.

**Requirements**: - Use faceting to create one subplot per region - Maintain consistent scales for valid comparison - Minimize repetitive axis labels - Remove unnecessary decorative elements - Ensure each panel is clearly labeled

```{r exercise-3}
# YOUR CODE HERE




```

**Expected Output**: A panel of four charts showing regional trends in a compact, comparable format.

------------------------------------------------------------------------

## Exercise 4: Dot Plot Instead of Bar Chart

**Task**: Replace a traditional bar chart with a dot plot (Cleveland dot plot) to show market share by product in 2023. Dot plots often have better data-ink ratios than bar charts.

**Requirements**: - Use points instead of bars - Include reference lines only where they add value - Sort products by market share for easier reading - Remove unnecessary elements

```{r exercise-4}
# YOUR CODE HERE




```

**Expected Output**: A dot plot that efficiently displays market share with minimal ink.

------------------------------------------------------------------------

## Exercise 5: Integrated Text and Graphics

**Task**: Create a scatter plot showing the relationship between units sold and customer satisfaction, with integrated text annotations for outliers or points of interest.

**Requirements**: - Add meaningful labels directly on or near important data points - Avoid overlapping text - Remove or minimize the legend - Use a minimal theme - Add a subtle reference line if it aids interpretation

```{r exercise-5}
# YOUR CODE HERE




```

**Expected Output**: A scatter plot where text and graphics work together to tell the story.

------------------------------------------------------------------------

## Exercise 6: Sparklines for Compact Trends

**Task**: Create sparkline-style visualizations showing revenue trends for each product across all years. Sparklines are small, word-sized graphics embedded in text or tables.

**Requirements**: - Create minimal line charts without axes - Remove all decorative elements - Focus on the shape of the trend - Make them small and compact

```{r exercise-6}
# YOUR CODE HERE
# Hint: Consider using facet_wrap with strip.text for labels
# and removing all theme elements for a true sparkline effect




```

**Expected Output**: Small, compact trend lines that show patterns at a glance.

------------------------------------------------------------------------

## Exercise 7: Avoid Misleading Scales

**Task**: Create two versions of a chart showing revenue growth: 1. A misleading version with a truncated y-axis 2. A proper version with appropriate scaling

**Requirements**: - First plot: Truncate y-axis to exaggerate differences - Second plot: Use full scale starting from zero - Add annotations explaining the difference - Discuss the implications in comments

```{r exercise-7}
# YOUR CODE HERE
# Version 1: Misleading (truncated axis)


# Version 2: Proper scaling


```

**Expected Output**: Two charts that demonstrate how scale choices affect perception.

------------------------------------------------------------------------

## Exercise 8: Range Frame Instead of Full Box

**Task**: Create a scatter plot using Tufte's "range frame" concept, where axis lines only extend to the data range rather than creating a full box.

**Requirements**: - Use `theme_tufte()` from ggthemes or create custom theme - Show only the data range on axes - Remove unnecessary grid lines - Focus attention on the data points

```{r exercise-8}
# YOUR CODE HERE




```

**Expected Output**: A plot with minimal framing that draws attention to the data.

------------------------------------------------------------------------

## Exercise 9: Data-Dense Table Graphic

**Task**: Create a table-graphic hybrid showing regional performance metrics, combining numerical precision with visual encoding.

**Requirements**: - Include small inline visualizations (e.g., bars or points) - Maintain alignment and readability - Remove table borders where they don't add value - Use subtle shading only if it improves readability

```{r exercise-9}
# YOUR CODE HERE
# Hint: Consider using geom_point or geom_segment within facets
# to create table-like layouts with visual elements




```

**Expected Output**: A data table enhanced with visual elements that aid quick comprehension.

------------------------------------------------------------------------

## Exercise 10: Comprehensive Dashboard

**Task**: Combine multiple Tufte-inspired visualizations into a single dashboard view showing the complete story of the sales data.

**Requirements**: - Use small multiples where appropriate - Maintain consistent styling across all plots - Integrate text annotations to guide interpretation - Remove all chartjunk - Ensure the overall layout is clean and uncluttered

```{r exercise-10}
# YOUR CODE HERE
# Hint: Consider using patchwork or cowplot packages for layout




```

**Expected Output**: A cohesive, information-dense dashboard that tells the complete data story.

------------------------------------------------------------------------

## Reflection Questions

After completing the exercises, consider:

1.  How does reducing non-data ink change the effectiveness of your visualizations?
2.  In what situations might you deviate from strict Tufte principles, and why?
3.  How do small multiples facilitate comparison better than overlaid elements?
4.  What challenges did you encounter when trying to maximize the data-ink ratio?

------------------------------------------------------------------------

## Additional Resources

- Tufte, E. R. (2001). *The Visual Display of Quantitative Information*
- Tufte, E. R. (1990). *Envisioning Information*
- `ggthemes` package documentation: https://jrnold.github.io/ggthemes/
- Cleveland, W. S. (1985). *The Elements of Graphing Data*

------------------------------------------------------------------------

## Summary

By completing these exercises, you've practiced creating visualizations that:

- Maximize information density while maintaining clarity
- Remove unnecessary decorative elements
- Use design choices that enhance rather than obscure the data
- Follow evidence-based principles of effective communication

Remember: the goal is not aesthetic minimalism for its own sake, but rather ensuring that every element of your visualization serves the purpose of communicating data effectively.