---
title: "Tufte Visualization Principles: Practice Exercises"
author: Daniel Wilkinson
format:
  html:
    toc: true
    code-fold: false
    theme: cosmo
    fig-width: 8
    fig-height: 6
execute:
  warning: false
  message: false

editor: visual
---

## Introduction

This document provides practice exercises in creating data visualizations that adhere to Edward Tufte's principles of effective graphical design. Tufte's principles emphasize:

1.  **Data-ink ratio**: Maximize the proportion of ink devoted to displaying actual data
2.  **Minimize chartjunk**: Remove unnecessary decorative elements
3.  **Show data variation, not design variation**: Let the data speak for itself
4.  **Use small multiples**: Facilitate comparison through repetition
5.  **Integrate text and graphics**: Labels should be close to the data they describe
6.  **Avoid misleading representations**: Use appropriate scales and proportions

## Dataset Description

I'll be working with a sample sales dataset (`sample_data.csv`) containing:

- **year**: Year of observation (2020-2023)
- **region**: Geographic region (North, South, East, West)
- **product**: Product category (Product_A, Product_B, Product_C)
- **revenue**: Sales revenue in dollars
- **units_sold**: Number of units sold
- **customer_satisfaction**: Customer satisfaction score (1-5 scale)
- **market_share**: Market share percentage

## Setup

```{r setup, message=FALSE, warning=FALSE}
# Load required libraries
library(tidyverse)
library(ggthemes)
library(gridExtra)
library(cowplot)
library(grid)

# Read the data
data <- read_csv("sample_data.csv")

# Preview the data structure
glimpse(data)
```

------------------------------------------------------------------------

## Exercise 1: High Data-Ink Ratio Line Chart

**Task**: Create a line chart showing revenue trends over time for each product, applying Tufte's principle of maximizing the data-ink ratio.

**Requirements**: - Remove unnecessary grid lines - Minimize or remove axis lines where appropriate - Use direct labels instead of a legend - Remove background shading - Keep only essential elements

```{r exercise-1}
# YOUR CODE HERE
data <- data |>
   mutate(product = gsub("_", " ", product)) 
data |>
  mutate(region = factor(region, levels = c("North", "East", "South", "West"))) |>
  ggplot(aes(x = year, y = revenue, color = product)) +
  geom_line() +
  facet_wrap(~ region) +
  geom_text(
    data = data |> filter(year == max(year)),
    aes(label = product),
    hjust = -0.1, size = 3
  ) +
      scale_colour_manual(values = c("Product A" = "grey20",
                                 "Product B" = "grey60",
                                 "Product C" = "#4a90a4")) +
  coord_cartesian(clip = "off") +
  theme(
    panel.grid       = element_blank(),
    panel.border     = element_blank(),
    axis.line        = element_line(colour = "black", linewidth = 0.3),
    panel.background = element_blank(),
    strip.background = element_blank(),
    legend.position  = "none",
    plot.margin      = margin(r = 60)
  ) +
  labs(x = "Year", y = "Revenue (USD)", 
       title = "Product C leads revenue in every region")

```

**Expected Output**: A clean line chart with minimal non-data ink, where product lines are clearly labeled and trends are easy to identify.

------------------------------------------------------------------------

## Exercise 2: Minimal Bar Chart

**Task**: Create a bar chart comparing total revenue by region for 2023, following minimalist design principles.

**Requirements**: - Remove grid lines - Remove chart borders - Use subtle colors or grayscale - Label bars directly (no legend needed) - Remove unnecessary axis elements

```{r exercise-2}
# YOUR CODE HERE
data |> 
  filter(year == 2023) |> 
  ggplot(aes(x = product, y = revenue)) + 
geom_col(fill = "grey80", colour = NA) +
geom_hline(yintercept = seq(50000, max(data$revenue), by = 50000),
           colour = "white", linewidth = 0.8) + 
  facet_wrap(~ region, labeller = labeller(region = toupper)) +
  geom_text(aes(label = scales::dollar(revenue)), vjust = -0.5, size = 3) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  theme(
    panel.grid       = element_blank(),
    panel.border     = element_blank(),
    panel.background = element_blank(),
    axis.line.x      = element_blank(),
    axis.ticks       = element_blank(),
    axis.text.y      = element_blank(),
    axis.title.y     = element_blank(),
    strip.background = element_blank(),
    legend.position  = "none",
    strip.text       = element_text(size = 11)
  ) +
  labs(x = NULL, title = "Product C generates most revenue in all regions, 2023")
```

**Expected Output**: A clean, uncluttered bar chart that focuses attention on the data values.

------------------------------------------------------------------------

## Exercise 3: Small Multiples for Comparison

**Task**: Create small multiples (faceted plots) to compare revenue trends across all four regions over time.

**Requirements**: - Use faceting to create one subplot per region - Maintain consistent scales for valid comparison - Minimize repetitive axis labels - Remove unnecessary decorative elements - Ensure each panel is clearly labeled

```{r exercise-3}
# YOUR CODE HERE

ggplot(data, aes(x = year, y = revenue, col = product)) + 
  geom_line() + 
  geom_point(size = 1.5, shape = 21, fill = "white", stroke = 0.4) +
  facet_wrap(~ region, labeller = labeller(region = toupper)) +
  scale_colour_manual(values = c("Product A" = "grey60", 
                                 "Product B" = "grey40", 
                                 "Product C" = "#4a90a4")) +

scale_y_continuous(
  limits = c(80000, 240000),
  breaks = c(80000, 120000, 160000, 200000, 240000),
  labels = scales::dollar,
  expand = c(0, 0)
)+
  theme(
    panel.grid       = element_blank(),
    panel.border     = element_blank(),
    axis.line        = element_line(colour = "black", linewidth = 0.3),
    panel.background = element_blank(),
    axis.line.x      = element_blank(),
    axis.ticks.x     = element_blank(),
    axis.title.y     = element_blank(),
    strip.background = element_blank(),
    legend.position  = "none",
    plot.margin      = margin(r = 10)
  ) +
theme(panel.spacing.x = unit(1, "cm"),
     panel.spacing.y = unit(1, "cm")) +
geom_text(
  data = data |> filter(year == max(year)),
  aes(label = product),
  hjust = 1, 
  vjust = -0.5,
  size = 2.5
) +
  labs(x = "", title = "Product C consistently generates the most revenue  (USD) in all regions")
```

**Expected Output**: A panel of four charts showing regional trends in a compact, comparable format.

------------------------------------------------------------------------

## Exercise 4: Dot Plot Instead of Bar Chart

**Task**: Replace a traditional bar chart with a dot plot (Cleveland dot plot) to show market share by product in 2023. Dot plots often have better data-ink ratios than bar charts.

**Requirements**: - Use points instead of bars - Include reference lines only where they add value - Sort products by market share for easier reading - Remove unnecessary elements

```{r exercise-4}
# YOUR CODE HERE
data |>
  filter(year == 2023) |>
  arrange(region, market_share) |>
  mutate(product = factor(product, levels = unique(product))) |>
  ggplot(aes(x = market_share, y = product)) +
  geom_segment(aes(x = min(market_share) -1 , xend = market_share, yend = product),
               colour = "grey80", linewidth = 0.3) +
  geom_point(size = 3, colour = "#4a90a4") +
geom_text(aes(label = paste0(round(market_share, 1), " %")),
          hjust = -0.5, size = 3) +
coord_cartesian(clip = "off") +
theme(plot.margin = margin(r = 20)) +
  facet_wrap(~ region, labeller = labeller(region = toupper)) +
  theme(
    panel.grid       = element_blank(),
    panel.border     = element_blank(),
    panel.background = element_blank(),
    strip.background = element_blank(),
    axis.line        = element_blank(),
    axis.ticks       = element_blank(),
    axis.title.x     = element_blank(),
    axis.text.x      = element_blank()) + 
theme(panel.spacing.y = unit(0.5, "cm")) +
  labs(y = NULL, title = "Product C leads market share in every region, 2023")
```

**Expected Output**: A dot plot that efficiently displays market share with minimal ink.

------------------------------------------------------------------------

## Exercise 5: Integrated Text and Graphics

**Task**: Create a scatter plot showing the relationship between units sold and customer satisfaction, with integrated text annotations for outliers or points of interest.

**Requirements**: - Add meaningful labels directly on or near important data points - Avoid overlapping text - Remove or minimize the legend - Use a minimal theme - Add a subtle reference line if it aids interpretation

```{r exercise-5}
ggplot(data, aes(x = units_sold, y = customer_satisfaction, col = product)) +
  geom_text(aes(label = substr(region, 1, 1)), size = 4, show.legend = FALSE) +
geom_text(
  data = data |> filter(year == min(year), region == "North"),
  aes(label = product),
  hjust = 0, nudge_x = 20, nudge_y = 0.35, size = 3, fontface = "bold"
)+
  facet_grid(year ~ .) +
  scale_colour_manual(values = c("Product A" = "grey20",
                                 "Product B" = "grey60",
                                 "Product C" = "#4a90a4")) +
  scale_y_continuous(breaks = seq(3, 5, by = 0.5)) +
  scale_x_continuous(breaks = c(min(data$units_sold), 400, 500, 600, max(data$units_sold))
) +
  coord_cartesian(ylim = c(3.5, 5), clip = "off") +
  theme(
    panel.grid       = element_blank(),
    panel.border     = element_blank(),
    panel.background = element_blank(),
    strip.background = element_blank(),
    axis.line        = element_blank(),
    axis.ticks       = element_blank(),
    legend.position  = "none",
    axis.text.y      = element_text(size = 8),
    strip.text.y     = element_text(angle = 0, size = 9),
    plot.margin      = margin(r = 60),
    panel.spacing.y  = unit(1, "cm"),
    plot.caption     = element_text(colour = "grey60"),
    plot.title   = element_text(size = 12),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12)
  ) +
  labs(x = "Number of Units Sold", y = "Customer Satisfaction (1-5)",
       title = "Number of Units Sold Correlates Positively with Customer Satisfaction Across All Years",
       caption = "The regions are indicated by 'N'='North', 'E'='East', 'S'='South', 'W'='West'.")
```

**Expected Output**: A scatter plot where text and graphics work together to tell the story.

------------------------------------------------------------------------

## Exercise 6: Sparklines for Compact Trends

**Task**: Create sparkline-style visualizations showing revenue trends for each product across all years. Sparklines are small, word-sized graphics embedded in text or tables.

**Requirements**: - Create minimal line charts without axes - Remove all decorative elements - Focus on the shape of the trend - Make them small and compact

```{r exercise-6}

data_spark <- data |>
  group_by(product, year) |>
  summarise(revenue = mean(revenue), .groups = "drop")

endpoints <- data_spark |>
  group_by(product) |>
  filter(revenue == min(revenue) | revenue == max(revenue))

ggplot(data_spark, aes(x = year, y = revenue)) +
  geom_line(linewidth = 0.5, colour = "grey20") +
  geom_point(data = endpoints, size = 1.5, colour = "grey20") +
  geom_text(
    data = endpoints |> filter(revenue == max(revenue)),
    aes(label = scales::dollar(revenue)),
    vjust = -0.8, size = 4
  ) +
  geom_text(
    data = endpoints |> filter(revenue == min(revenue)),
    aes(label = scales::dollar(revenue)),
    vjust = 1.8, size = 4
  ) +
  facet_wrap(~ product, nrow = 1) +
  scale_x_continuous(limits = c(2019.5, 2024), expand = c(0, 0)) +
  scale_y_continuous(limits = c(min(data_spark$revenue), max(data_spark$revenue))) +
  coord_cartesian(clip = "off") +
  theme_void() +
  theme(
    strip.text    = element_text(size = 10, colour = "grey40"),
    panel.spacing = unit(0.5, "cm"),
    plot.margin   = margin(10, 15, 10, 10),
      plot.caption     = element_text(colour = "grey60")
  ) +
  
labs(caption = "Values shown are averages across all four regions")
```

**Expected Output**: Small, compact trend lines that show patterns at a glance.

------------------------------------------------------------------------

## Exercise 7: Avoid Misleading Scales

**Task**: Create two versions of a chart showing revenue growth: 1. A misleading version with a truncated y-axis 2. A proper version with appropriate scaling

**Requirements**: - First plot: Truncate y-axis to exaggerate differences - Second plot: Use full scale starting from zero - Add annotations explaining the difference - Discuss the implications in comments

```{r exercise-7}

data_2023 <- data |> filter(year == 2023) |>
  group_by(product) |>
  summarise(revenue = mean(revenue), .groups = "drop")

p1 <- ggplot(data_2023, aes(x = product, y = revenue)) +
  geom_col(fill = "grey80") +
  scale_y_continuous(labels = scales::dollar, expand = c(0, 0)) +
  coord_cartesian(ylim = c(130000, 210000)) +
  annotate("text", x = 1, y = 205000, label = "Truncated axis — \ndifferences look huge",
           size = 4, colour = "#a32d2d", hjust = 0) +
  theme(panel.grid = element_blank(), panel.background = element_blank(),
        axis.ticks.x = element_blank(), panel.border = element_blank(),
        axis.line.y = element_line(linewidth = 0.3)) +
  labs(x = NULL, y = NULL, title = "Misleading: starts at $130,000", size=11)

p2 <- ggplot(data_2023, aes(x = product, y = revenue)) +
  geom_col(fill = "grey80") +
  scale_y_continuous(labels = scales::dollar, expand = c(0, 0)) +
  coord_cartesian(ylim = c(0, 220000)) +
  annotate("text", x = 1, y = 210000, label = "Full scale —\n true proportions visible",
           size = 4, colour = "#3b6d11", hjust = 0) +
  theme(panel.grid = element_blank(), panel.background = element_blank(),
        axis.ticks.x = element_blank(), panel.border = element_blank(),
        axis.line.y = element_line(linewidth = 0.3)) +
  labs(x = NULL, y = NULL, title = "Honest: starts at zero")

title <- ggdraw() + 
  draw_label("The same data, two very different stories", 
             fontface = "plain", size = 11, colour = "grey20")

plot_grid(title, plot_grid(p1, p2, ncol = 2), 
          ncol = 1, rel_heights = c(0.1, 1))
```

**Expected Output**: Two charts that demonstrate how scale choices affect perception.

------------------------------------------------------------------------

## Exercise 8: Range Frame Instead of Full Box

**Task**: Create a scatter plot using Tufte's "range frame" concept, where axis lines only extend to the data range rather than creating a full box.

**Requirements**: - Use `theme_tufte()` from ggthemes or create custom theme - Show only the data range on axes - Remove unnecessary grid lines - Focus attention on the data points

```{r exercise-8}

ggplot(data, aes(x = units_sold, y = revenue)) +
  geom_point(colour = "grey40", size = 2, alpha = 0.7) +
  geom_rangeframe() +
  scale_y_continuous(
    breaks = c(min(data$revenue), max(data$revenue)),
    labels = scales::dollar
  ) +
  scale_x_continuous(
    breaks = c(min(data$units_sold), max(data$units_sold)),
    labels = scales::comma
  ) +
    geom_smooth(method = "lm", se = FALSE, 
            linewidth = 0.3, colour = "grey60") +
  theme_tufte() +
  labs(x = "Units sold", y = "",
       title = "Revenue scales (USD) with units sold")
```

**Expected Output**: A plot with minimal framing that draws attention to the data.

------------------------------------------------------------------------

## Exercise 9: Data-Dense Table Graphic

**Task**: Create a table-graphic hybrid showing regional performance metrics, combining numerical precision with visual encoding.

**Requirements**: - Include small inline visualizations (e.g., bars or points) - Maintain alignment and readability - Remove table borders where they don't add value - Use subtle shading only if it improves readability

```{r exercise-9}

data_table <- data |>
  filter(year == 2023) |>
  group_by(region) |>
  summarise(
    revenue      = mean(revenue),
    units_sold   = mean(units_sold),
    satisfaction = mean(customer_satisfaction),
    market_share = mean(market_share),
    .groups = "drop"
  ) |>
  arrange(revenue) |>
  mutate(region = factor(region, levels = region))


p_text <- ggplot(data_table, aes(y = region)) +
  geom_text(aes(x = 0, label = region), hjust = 0, size = 3, fontface = "bold") +
  geom_text(aes(x = 1, label = scales::dollar(revenue, accuracy = 1)), hjust = 1, size = 3) +
  geom_text(aes(x = 1.4, label = round(units_sold, 0)), hjust = 1, size = 3) +
  geom_text(aes(x = 1.8, label = round(satisfaction, 1)), hjust = 1, size = 3) +
  annotate("text", x = 0,   y = 4.7, label = "Region",       hjust = 0, size = 2.8, colour = "grey40", fontface = "bold") +
  annotate("text", x = 1,   y = 4.7, label = "Avg revenue",  hjust = 1, size = 2.8, colour = "grey40", fontface = "bold") +
  annotate("text", x = 1.4, y = 4.7, label = "Units",        hjust = 1, size = 2.8, colour = "grey40", fontface = "bold") +
  annotate("text", x = 1.8, y = 4.7, label = "Satisfaction", hjust = 1, size = 2.8, colour = "grey40", fontface = "bold") +
  annotate("segment", x = 0, xend = 2, y = 4.5, yend = 4.5,
           colour = "grey60", linewidth = 0.3) +
  scale_x_continuous(limits = c(0, 2)) +
  scale_y_discrete(expand = expansion(add = c(0.5, 1))) +
  theme_void() +
  theme(plot.margin = margin(0, 0, 0, 10))

p_bars <- ggplot(data_table, aes(y = region, x = market_share)) +
  geom_col(fill = "#4a90a4", width = 0.3, alpha = 0.6) +
  geom_text(aes(label = paste0(round(market_share, 1), "%")),
            hjust = -0.1, size = 2.5, colour = "grey20") +
  annotate("text", x = 0, y = 4.7, label = "Market share",
           hjust = 0, size = 2.8, colour = "grey40", fontface = "bold") +
  annotate("segment", x = 0, xend = 30, y = 4.5, yend = 4.5,
           colour = "grey60", linewidth = 0.3) +
  scale_x_continuous(limits = c(0, 30), expand = c(0, 0)) +
  scale_y_discrete(expand = expansion(add = c(0.5, 1))) +
  theme_void() +
  theme(plot.margin = margin(0, 10, 0, 0))

grid.arrange(p_text, p_bars, ncol = 2, widths = c(2.5, 1),
             top = "Regional performance · 2023")
```

**Expected Output**: A data table enhanced with visual elements that aid quick comprehension.

------------------------------------------------------------------------

## Exercise 10: Comprehensive Dashboard

**Task**: Combine multiple Tufte-inspired visualizations into a single dashboard view showing the complete story of the sales data.

**Requirements**: - Use small multiples where appropriate - Maintain consistent styling across all plots - Integrate text annotations to guide interpretation - Remove all chartjunk - Ensure the overall layout is clean and uncluttered

```{r exercise-10}

# panel 1 — revenue trend by product (sparkline style)
data_spark <- data |>
  group_by(product, year) |>
  summarise(revenue = mean(revenue), .groups = "drop")

endpoints <- data_spark |>
  group_by(product) |>
  filter(revenue == min(revenue) | revenue == max(revenue))

p1 <- ggplot(data_spark, aes(x = year, y = revenue, colour = product)) +
  geom_line(linewidth = 0.6) +
  geom_point(data = endpoints, size = 1.5) +
  geom_text(
    data = endpoints |> filter(revenue == max(revenue)),
    aes(label = scales::dollar(revenue, accuracy = 1)),
    vjust = -0.8, size = 2.2
  ) +
  geom_text(
    data = data_spark |> filter(year == max(year)),
    aes(label = product),
    hjust = -0.1, size = 2.5
  ) +
  scale_colour_manual(values = c("Product A" = "grey60",
                                 "Product B" = "grey40",
                                 "Product C" = "#4a90a4")) +
  scale_x_continuous(limits = c(2020, 2025), expand = c(0, 0)) +
  scale_y_continuous(labels = scales::dollar) +
  coord_cartesian(clip = "off") +
  theme(
    panel.grid       = element_blank(),
    panel.background = element_blank(),
    panel.border     = element_blank(),
    axis.line.y      = element_line(linewidth = 0.3, colour = "grey60"),
    axis.ticks.x     = element_blank(),
    axis.title       = element_blank(),
    legend.position  = "none",
    plot.margin      = margin(5, 60, 5, 5),
    plot.title       = element_text(size = 9, colour = "grey40")
  ) +
  labs(title = "1. Product C leads revenue growth")

# panel 2 — dot plot market share 2023
data_2023 <- data |>
  filter(year == 2023) |>
  group_by(product) |>
  summarise(market_share = mean(market_share), .groups = "drop") |>
  arrange(market_share) |>
  mutate(product = factor(product, levels = product))

p2 <- ggplot(data_2023, aes(x = market_share, y = product)) +
  geom_segment(aes(x = min(market_share) - 1, xend = market_share, yend = product),
               colour = "grey80", linewidth = 0.3) +
  geom_point(size = 3, colour = "#4a90a4") +
  geom_text(aes(label = paste0(round(market_share, 1), "%")),
            hjust = -0.4, size = 2.5) +
  coord_cartesian(clip = "off") +
  theme(
    panel.grid       = element_blank(),
    panel.background = element_blank(),
    panel.border     = element_blank(),
    axis.line        = element_blank(),
    axis.ticks       = element_blank(),
    axis.title       = element_blank(),
    axis.text.x      = element_blank(),
    legend.position  = "none",
    plot.margin      = margin(5, 40, 5, 5),
    plot.title       = element_text(size = 9, colour = "grey40")
  ) +
  labs(title = "2. Market share by product, 2023")

# panel 3 — units sold vs satisfaction
p3 <- ggplot(data, aes(x = units_sold, y = customer_satisfaction, colour = product)) +
  geom_text(aes(label = substr(region, 1, 1)), size = 3) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.3, colour = "grey60") +
  scale_colour_manual(values = c("Product A" = "grey60",
                                 "Product B" = "grey40",
                                 "Product C" = "#4a90a4")) +
  scale_y_continuous(limits = c(3.5, 5), breaks = seq(3.5, 5, 0.5)) +
  theme(
    panel.grid       = element_blank(),
    panel.background = element_blank(),
    panel.border     = element_blank(),
    axis.line        = element_line(linewidth = 0.3, colour = "grey60"),
    axis.ticks       = element_blank(),
    axis.title       = element_text(size = 8, colour = "grey40"),
    legend.position  = "none",
    plot.margin      = margin(5, 5, 5, 5),
    plot.title       = element_text(size = 9, colour = "grey40")
  ) +
  labs(x = "Units sold", y = "Satisfaction",
       title = "3. More units sold, higher satisfaction")

# panel 4 — regional revenue table
data_region <- data |>
  filter(year == 2023) |>
  group_by(region) |>
  summarise(
    revenue      = mean(revenue),
    units_sold   = mean(units_sold),
    satisfaction = mean(customer_satisfaction),
    market_share = mean(market_share),
    .groups = "drop"
  ) |>
  arrange(revenue) |>
  mutate(region = factor(region, levels = region))

p4 <- ggplot(data_region, aes(y = region)) +
  geom_text(aes(x = 0,   label = region),                               hjust = 0, size = 2.8, fontface = "bold") +
  geom_text(aes(x = 1,   label = scales::dollar(revenue, accuracy=1)),  hjust = 1, size = 2.8) +
  geom_text(aes(x = 1.4, label = round(units_sold, 0)),                 hjust = 1, size = 2.8) +
  geom_text(aes(x = 1.8, label = round(satisfaction, 1)),               hjust = 1, size = 2.8) +
  geom_text(aes(x = 2.2, label = paste0(round(market_share,1), "%")),   hjust = 1, size = 2.8) +
  annotate("text", x = 0,   y = 4.7, label = "Region",      hjust = 0, size = 2.5, colour = "grey40", fontface = "bold") +
  annotate("text", x = 1,   y = 4.7, label = "Revenue",     hjust = 1, size = 2.5, colour = "grey40", fontface = "bold") +
  annotate("text", x = 1.4, y = 4.7, label = "Units",       hjust = 1, size = 2.5, colour = "grey40", fontface = "bold") +
  annotate("text", x = 1.8, y = 4.7, label = "Satisfaction",hjust = 1, size = 2.5, colour = "grey40", fontface = "bold") +
  annotate("text", x = 2.2, y = 4.7, label = "Mkt share",   hjust = 1, size = 2.5, colour = "grey40", fontface = "bold") +
  annotate("segment", x = 0, xend = 2.3, y = 4.5, yend = 4.5,
           colour = "grey60", linewidth = 0.3) +
  scale_x_continuous(limits = c(0, 2.3)) +
  scale_y_discrete(expand = expansion(add = c(0.5, 1))) +
  theme_void() +
  theme(
    plot.margin = margin(5, 5, 5, 5),
    plot.title  = element_text(size = 9, colour = "grey40")
  ) +
  labs(title = "4. Regional performance, 2023")

# combine
grid.arrange(
  p1, p2, p3, p4,
  ncol = 2,
  top = grid::textGrob(
    "Product C dominates — but satisfaction and volume rise together across all products",
    gp = grid::gpar(fontsize = 11, col = "grey20")
  )
)

```

**Expected Output**: A cohesive, information-dense dashboard that tells the complete data story.

------------------------------------------------------------------------

## Reflection Questions

1.  How does reducing non-data ink change the effectiveness of your visualizations?
    - reduced visual distractions
    - quicker to determine the key message
2.  In what situations might you deviate from strict Tufte principles, and why?
    - when memorability or engagement (through e.g. colour) are priorities
    - removing gridlines, for example, can stop you from seeing exact values
    - to add contextual elements like trend lines
3.  How do small multiples facilitate comparison better than overlaid elements?
    - reduce clutter and overlapping
    - it is easier to isolate patterns in the data
    - enforces consistent scales
    - highlight outliers and trends
4.  What challenges did you encounter when trying to maximize the data-ink ratio?
    - it requires many manual code adjustments, especially to the theme. This means that plotting according to tufte principles cannot be easily automated because of the difference in meaning/intention even between two plots of the same type and data types.

------------------------------------------------------------------------

## Additional Resources

- Tufte, E. R. (2001). *The Visual Display of Quantitative Information*
- Tufte, E. R. (1990). *Envisioning Information*
- `ggthemes` package documentation: https://jrnold.github.io/ggthemes/
- Cleveland, W. S. (1985). *The Elements of Graphing Data*

------------------------------------------------------------------------

## Summary

By completing these exercises, you've practiced creating visualizations that:

- Maximize information density while maintaining clarity
- Remove unnecessary decorative elements
- Use design choices that enhance rather than obscure the data
- Follow evidence-based principles of effective communication

Remember: the goal is not aesthetic minimalism for its own sake, but rather ensuring that every element of your visualization serves the purpose of communicating data effectively.