---
title: "Environmental Data Analysis"
author: "plotting_in_r"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

## Overview

This document analyses a sample environmental dataset containing measurements
collected across multiple national park locations. The dataset includes
**numerical**, **string**, and **categorical** variables.

## Setup

Load required packages. These are managed via `renv` — run `renv::restore()`
once to install them before rendering this document.

```{r}
#| label: setup

library(ggplot2)
library(dplyr)
library(readr)
library(tidyr)
library(here)
```

## Load Data

```{r}
#| label: load-data

env_data <- read_csv(
  here::here("data", "environmental_data.csv"),
  col_types = cols(
    sample_id       = col_character(),
    date            = col_date(format = "%Y-%m-%d"),
    location        = col_character(),
    observer_name   = col_character(),
    season          = col_factor(levels = c("Spring", "Summer", "Autumn", "Winter")),
    weather_condition = col_factor(),
    land_use_type   = col_factor(),
    temperature_c   = col_double(),
    humidity_pct    = col_double(),
    wind_speed_kmh  = col_double(),
    precipitation_mm = col_double(),
    air_quality_index = col_integer(),
    uv_index        = col_double(),
    visibility_km   = col_double()
  )
)

glimpse(env_data)
```

### Variable Overview

| Variable | Type | Description |
|---|---|---|
| `sample_id` | String | Unique record identifier |
| `date` | String / Date | Date of observation |
| `location` | String | National park name |
| `observer_name` | String | Name of field observer |
| `season` | Categorical | Season of observation |
| `weather_condition` | Categorical | Observed weather type |
| `land_use_type` | Categorical | Ecosystem / land-use classification |
| `temperature_c` | Numerical | Air temperature (°C) |
| `humidity_pct` | Numerical | Relative humidity (%) |
| `wind_speed_kmh` | Numerical | Wind speed (km/h) |
| `precipitation_mm` | Numerical | Precipitation (mm) |
| `air_quality_index` | Numerical | AQI score |
| `uv_index` | Numerical | UV radiation index |
| `visibility_km` | Numerical | Atmospheric visibility (km) |

## Summary Statistics

```{r}
#| label: summary-stats

numerical_vars <- c(
  "temperature_c", "humidity_pct", "wind_speed_kmh",
  "precipitation_mm", "air_quality_index", "uv_index", "visibility_km"
)

env_data |>
  select(all_of(numerical_vars)) |>
  summary()
```

### Observations per Season

```{r}
#| label: season-counts

env_data |>
  count(season, name = "n_observations") |>
  arrange(season)
```

## Visualisations

### Temperature Distribution by Season

```{r}
#| label: fig-temp-season
#| fig-cap: "Distribution of temperature (°C) for each season."

ggplot(env_data, aes(x = season, y = temperature_c, fill = season)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Temperature Distribution by Season",
    x = "Season",
    y = "Temperature (°C)"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

### Temperature vs. Humidity

```{r}
#| label: fig-temp-humidity
#| fig-cap: "Scatter plot of temperature against humidity coloured by season."

ggplot(env_data, aes(x = temperature_c, y = humidity_pct, colour = season)) +
  geom_point(alpha = 0.7, size = 2.5) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.8) +
  scale_colour_brewer(palette = "Set2") +
  labs(
    title = "Temperature vs. Humidity",
    x = "Temperature (°C)",
    y = "Humidity (%)",
    colour = "Season"
  ) +
  theme_minimal(base_size = 13) +
  facet_wrap(~season)
```

### Air Quality Index by Land Use Type

```{r}
#| label: fig-aqi-landuse
#| fig-cap: "Air Quality Index distribution across land use categories."

env_data |>
  group_by(land_use_type) |>
  summarise(
    mean_aqi  = mean(air_quality_index),
    se_aqi    = sd(air_quality_index) / sqrt(n()),
    n         = n()
  ) |>
  ggplot(aes(x = reorder(land_use_type, mean_aqi), y = mean_aqi, fill = land_use_type)) +
  geom_col(alpha = 0.8) +
  geom_errorbar(aes(ymin = mean_aqi - se_aqi, ymax = mean_aqi + se_aqi), width = 0.3) +
  coord_flip() +
  scale_fill_brewer(palette = "Paired") +
  labs(
    title = "Mean Air Quality Index by Land Use Type",
    x = "Land Use Type",
    y = "Mean AQI",
    caption = "Error bars show ±1 SE"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

### Precipitation Heatmap: Season × Weather Condition

```{r}
#| label: fig-precip-heatmap
#| fig-cap: "Mean precipitation (mm) by season and weather condition."

env_data |>
  group_by(season, weather_condition) |>
  summarise(mean_precip = mean(precipitation_mm), .groups = "drop") |>
  ggplot(aes(x = season, y = weather_condition, fill = mean_precip)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  scale_fill_viridis_c(option = "plasma", name = "Precipitation\n(mm)") +
  labs(
    title = "Mean Precipitation by Season and Weather Condition",
    x = "Season",
    y = "Weather Condition"
  ) +
  theme_minimal(base_size = 12)
```

### Wind Speed Distribution

```{r}
#| label: fig-wind-dist
#| fig-cap: "Histogram of wind speed coloured by weather condition."

ggplot(env_data, aes(x = wind_speed_kmh, fill = weather_condition)) +
  geom_histogram(binwidth = 5, colour = "white", alpha = 0.85) +
  scale_fill_brewer(palette = "Set3") +
  labs(
    title = "Wind Speed Distribution by Weather Condition",
    x = "Wind Speed (km/h)",
    y = "Count",
    fill = "Weather Condition"
  ) +
  theme_minimal(base_size = 13)
```

## Correlation Analysis

```{r}
#| label: fig-correlation
#| fig-cap: "Pairwise correlation matrix for numerical variables."

cor_matrix <- env_data |>
  select(all_of(numerical_vars)) |>
  cor(use = "complete.obs")

cor_long <- as.data.frame(as.table(cor_matrix)) |>
  rename(var1 = Var1, var2 = Var2, correlation = Freq)

ggplot(cor_long, aes(x = var1, y = var2, fill = correlation)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = round(correlation, 2)), size = 3) +
  scale_fill_gradient2(
    low = "#0571b0", mid = "white", high = "#ca0020",
    midpoint = 0, limits = c(-1, 1), name = "Correlation"
  ) +
  labs(
    title = "Correlation Matrix — Numerical Variables",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Session Information

```{r}
#| label: session-info

sessionInfo()
```
